Linking microbial communities to ecosystem functions: what we can learn from genotype-phenotype mapping in organisms

### 2019-07-19 manuscript/main.tex

\documentclass{article}
%\usepackage[utf8]{inputenc}
%\usepackage{amsmath}
%\usepackage{amsfonts}
%\usepackage{amssymb}
%\usepackage{textalpha}
\usepackage{natbib}

@@

Is variation in microbial community
structure associated with differences in ecosystem function across space,
time, and environments?
\section{Methods}

@@

Next, we used a statistical approach common in population genetics genome-wide
association studies. We fit variance component mixed models from the
\texttt{varComp} package to test whether variation in microbial
community composition expains a significant portion of the variation in
ecosystem function \citep{qu2013}. First, we fit an intercept-only model with one random effect
for community similarity calculated from Bray-Curtis dissimilarity for abundance
data and the Jaccard index for presence/absance data. We tested the significance
of these components using a linear score test by comparing nested models with and
without the microbial community variance component \citep{qu2013}. We then estimated
the community heritability of this function by calculating the degree of
variation in the rate of ecosystem function due to variation in community
composition between ecosystems including standard errors on that estimate 
using the \texttt{h2G} function from the
\texttt{gap} package. We then tested whether microbial community composition explains a
significant portion of the variation in ecosystem function after accounting for
variation in the environment and accounting for the spatial distance between
samples. We did this by fitting an intercept-only model with variance components
for environmental similarity from euclidean distance of a matrix of
environmental covariates and spatial similarity
also using euclidean distance from the latitude and longitude coordinates where
the samples were collected. We then used a linear score test to compare
nested models with and without the microbial community variance component to
determine whether community composition adds any variance explained after
accounting for spatial and environmental covariates. We then estimated the
variance explained by each component (environment, geography, and community
composition) with standard errors of the estimates using the \texttt{h2GE}
function.

@@

To demonstrate the ideas presented in the Introduction, we re-analyzed data from
a forthcoming publication \citep{meyer2019}. With these data, we demonstrate the
importance of selecting the appropriate population to sample from. We are
precise about the variation we are measuring. Specifically, we measured the rate
of a specific microbial pathway. In order to minimize environmenal variation, we
brought intact soil cores into the lab to incubate them under similar
temperature and methane concentration. We also statistically controlled for
environmental covariates that we couldn't isolate in the lab including soil
carbon and nitrogen, soil bulk density, and soil moisture content. Finally, we
surveyed the microbial community agnostically for microbial markers of ecosystem
function.

We found that for two pathways of methane oxidation in

\begin{itemize}
  \item For low-affinity and high-affinity methane oxidation in Gabon and
    methane production in Brazil, community composition explains a significant
    portion of the variation in ecosystem function. (linear score test or
    likelihood ratio test)
  \item On its own, the portion of `community heritability' varies by function.
    \textasciitilde99\% (SE = 20\%) for high affinity, \textasciitilde61\% (SE =
    39\%)for low affinity, etc. 
  \item Geographic proximity: high affinity 0.85 (0.10), low affinity n.s. 0.34
    (0.31)
  \item Environmental similarity: high affinity 0.61 (0.23), low affinity 0.61 (0.36) 
  \item Some of these become n.s. when combined suggesting they explain an
    overlapping portion of variation. Perhaps this means we should fit the
    variance component model with just the community similarity matrix knowing
    that that will capture the important components of historical effects and
    selection.
  \item Function varies strongly by ecosystem type: in Brazil, pasture has
    higher methane fluxes than primary or secondary forest; in Gabon, peatlands
    and wetlands have higher low affinity while plantations and forests have
    higher high affinity.
  \item Possible next step: fit a variance component model with community
    similarity as the random effect and taxon abundance as the fixed effect and
    collect p.values for each taxon (though this may return none). Or include
    all three variance components. Adjust for false discovery rate to get
    significant taxa.
\end{itemize}

@@

In this paper, we argue that the traditional approach to microbial BEF research
ignores important understandings about the nature of ecological processes in
determining the distribution of microorganisms. The marker gene hypothesis has
proven unsuccesful in predicting the rates of ecosystem functions from the
abundance of functional genes and in our example study was not significant in one case and
was significant in another, though explained only a small portion of the variation.
Similarly, diversity metrics from the entire Prokrayotic community were not
explanatory. These results agree with recent reviews of this literature and
demonstrate that these approaches are not a fruitful avenue for elucidating
microbial structure-function connections. After identifying significant taxa,
it is notable that none of the significant taxa fall into expected functional
groups. For example, none of the taxa identified in this analysis were
methanotrophs.

@@

@@

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage{subfiles}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{graphicx}
\graphicspath{{../Figures/}}

@@

Is variation in microbial community
structure associated with differences in ecosystem function across space,
time, and environments?

\section{Methods}

@@

Next, we used a statistical approach common in population genetics genome-wide
association studies. We fit variance component mixed models from the
\texttt{varComp} package to test whether variation in microbial
community composition expains a significant portion of the variation in
ecosystem function \citep{qu2013}. First, we fit an intercept-only model with
each random effect in all combinations. Variance components included deographic location was coded as a factor, environmental
similarity from euclidean distance of four edaphic variables including total
nitrogen, total carbon, moisture content, and bulk density, and
community similarity calculated from Bray-Curtis dissimilarity for abundance
data and the Jaccard index for presence/absance data. We tested the significance
of these components using a linear score test by comparing nested models with and
without each variance component \citep{qu2013}. We then estimated
the percent of variation in ecosystem function explained by each component 
including standard errors on that estimate 
using the \texttt{h2G} function from the
\texttt{gap} package. We then tested whether microbial community composition explains a
significant portion of the variation in ecosystem function after accounting for
variation in the environment and accounting for the spatial distance between
samples. We did this by fitting an intercept-only model with variance components
for environmental similarity from euclidean distance of a matrix of
environmental covariates and spatial similarity
also using euclidean distance from the latitude and longitude coordinates where
the samples were collected. We then used a linear score test to compare
nested models with and without the microbial community variance component to
determine whether community composition adds any variance explained after
accounting for spatial and environmental covariates. We then estimated the
variance explained by each component (environment, geography, and community
composition) with standard errors of the estimates using the \texttt{h2GE}
function.

@@

\includegraphics[width=0.5\textwidth]{lowk_pmoa}
\includegraphics[width=0.5\textwidth]{lowk_pc1}
\includegraphics[width=0.5\textwidth]{lowk_rich}
\includegraphics[width=0.5\textwidth]{lowk_shan}

All three variance components are significant individually. Community
significant in the community + environment model and geography
significant in the geography + environment model. Table \ref{tab:pvalues}

\subfile{../tables/var_comp_pvalues}

Including different covariates removes different numbers of
significant taxa. Some covariates add taxa that were not previously
identified as significant. Table \ref{tab:n_taxa}

\subfile{../tables/all_taxa_n}

After controlling for geographic proximity, environmental similarity, and
similarity in community composition, four taxa were significantly correlated
with the rate of methane oxidation. None of these four taxa are known methane
oxidizers or methane producers. Table  \ref{tab:sig_taxa}

\subfile{../tables/all_taxa_sig}

@@

In this paper, we argue that the traditional approach to microbial BEF research
ignores important understandings about the nature of ecological processes in
determining the distribution of microorganisms. The marker gene hypothesis has
proven unsuccesful in predicting the rates of ecosystem functions from the
abundance of functional genes and in our example study was not significant.
Similarly, diversity metrics from the entire Prokaryotic community were not
explanatory. These results agree with recent reviews of this literature and
demonstrate that these approaches are not a fruitful avenue for elucidating
microbial structure-function connections.

@@

There are two ways to interpret an association between taxon abundance and rate
of function. One is a statistical co-occurrence and the other is a biological
interpetation, for example pH affects function and pH affects taxon or pH
affects taxon, which affects function.
A significant association between a taxon and the rate of ecosystem function
could be explained a number of ways. One possibility is that the taxon is
statistically related, because it is causally connected to the function. This
could be direct, for example an organism that consumes methane, or indirect, for
example an organism that regulates substrates necessary for consumers of
methane. In either of these cases, this would represent a taxon that could be
useful as a biomarker of function or an organism to investigate to understand
the mechanistic relationship between communities and function. Alternativelly, a significant
association could occur due to admixture within a community. In a community of
microorganisms, any organism that tends to be in high abundance under the same
conditions as methane oxidizers will be correlated with methane oxidation even
if it has no mechanistic relationship with the rate of methane flux. In fact,
any taxa in high abundance in highly methanotrophic communities will be
statistically related to function.

In this variance component modeling framework, many taxa that are significantly
associated with function become non-significant after accounting for different
covariates. This could be explained by different relationships between the
environment, community assembly processes, and the rate of ecosystem function.
For example, a taxon significantly correlated with function that becomes
non-significant after controlling for the underlying environmental variation
could be acting as a conduit through which the environment alters function.
Alternatively, if a taxon becomes non-significant after controlling for
geographic proximity, this taxon could be related to function due to historical
effects such as past environmental conditions or through dispersal from a nearby
patch. Alternatively, geography could represent unmeasured environmental
variables that change over distance. Finally, if incorporating community
similarity makes a taxon non-significant, then

None of the significant taxa identified by the full variance component model 
fall into expected functional groups. For example, none of the taxa identified
in this analysis were methanotrophs or methanogens.

Broad vs narrow processes: Some genetic diseases such as Parkinson's 
are controlled by a single locus. However, in most GWAS studies the phenotype of
interest is a result of many loci of small effect (as well as environmental
factors) such as human height and LDL cholesterol. For ecosystem functions that
consist of physiologically or phylogenetically "narrow" process (sensu
\cite{schimel1995b}), it's possible that a single marker gene could control the
rate of a process. For example, methane flux in permafrost in Sweden may be
controlled by a single taxon \citep{mccalley2014}. However, most microbial ecosystem functions are likely the result
of many taxa interacting directly or indirectly with function.
Trying to identify many taxa of small effect is made easier by constraining that
variation by defining your population in a way tha makes them similar: 
similar ecosystem, soil type, and abiotic conditions.

With these data, we demonstrate the
importance of selecting the appropriate population to sample from. We are
precise about the variation we are measuring. Specifically, we measured the rate
of a specific microbial pathway. In order to minimize environmenal variation, we
brought intact soil cores into the lab to incubate them under similar
temperature and methane concentration. We also statistically controlled for
environmental covariates that we couldn't isolate in the lab including soil
carbon and nitrogen, soil bulk density, and soil moisture content. Finally, we
surveyed the microbial community agnostically for microbial markers of ecosystem
function.

@@

\subsection{Caveats}

Future studies could improve on the approach illustrated here in a number of
ways. First, GWAS studies rely on high power to detect significant associations
between phenotype and SNPs of small effect. This study consistent of 44 samples,
which is typical for microbial ecology sudies, but would benefit from a large
sampling of ecosystems and variation in both community composition and function. 
Second, this exploratory study sampled across
a diversity of wetland and upland ecosystems in the Congo basin. However, upland
ecosystems such as forests are known to have the greatest rates of high-affinity
methane oxidation. Identifying connections between variation in community
composition and variation in ecosystem function would benefit from sampling a
population of ecosystems that is known to exhibit variation in the rate of
the target function. Third, it is important to define a narrow pathway
definition. Functions such as low-affinity methane oxidation and bulk methane
production are relatively broad pathways and as such exhibit no significant
correlation with individual taxa \citep{meyer2019}. Here we have measured
high-affinity methane oxidation, which is performed by relatively few taxa.
Functions such as this are more phylogenetically conserved and as such are more
likely to have significant connections between community and function. Fourth,
these environmental variables measured may or may not be the correct ones to
measure to accurately account for the environmental effect on function.
Moisture, structure, total nitrogen, and total carbon are relatively course
measurements of environmental variation. Including more environmental variables
and variables known to influence the rate of the function of interest would more
accurately estimate the effect of the environment of communities and function.
This could also be applied to the geographic location of samples. Our study
sites were clustered around three locations that are far apart and so didn't
fully capturing the geographic variation underlying these ecosystems. Sampling
in a grid or logarithmic pattern might more suitable capture variation across
the landscape.
